---
title: "Multi-temporal urban sprawl detection and growth-pattern analysis"
subtitle: "Research report — methods, results, limitations, and research gaps"
author: "Urban Sprawl R pipeline (Phases 1–7)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    self_contained: true
    md_extensions: "+raw_html"
    toc: true
    toc_depth: 3
    toc_float: false
    theme: flatly
    highlight: tango
    number_sections: true
    number_section_depth: 3
    fig_caption: true
    df_print: kable
    code_folding: hide
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    keep_tex: false
    latex_engine: xelatex
    includes:
      in_header: latex_preamble.tex
params:
  project_root: "."
  study_name: "Study area"
  outputs_dir: "outputs"
  outputs_root: null
---

```{r setup, include=FALSE}
# render() with no output_format uses the *first* YAML output. html_document is listed first so
# output_file = "...html" is not filled with PDF binary. For PDF: output_format = "pdf_document".
`%||%` <- function(x, y) if (is.null(x)) y else x
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 144
)
root_param <- normalizePath(params$project_root, winslash = "/", mustWork = FALSE)
if (!nzchar(root_param) || !dir.exists(root_param)) {
  stop("params$project_root must exist: ", params$project_root, call. = FALSE)
}
outd <- params$outputs_dir %||% "outputs"
# Project root from this file: docs/urban_sprawl_report.Rmd -> parent of docs/ (ignores wrong getwd).
infer_root_from_rmd <- tryCatch(
  {
    rd <- knitr::current_input(dir = TRUE)
    if (is.null(rd) || !nzchar(rd)) {
      return(NULL)
    }
    doc_dir <- normalizePath(rd, winslash = "/", mustWork = FALSE)
    cand <- normalizePath(file.path(doc_dir, ".."), winslash = "/", mustWork = FALSE)
    yml <- file.path(cand, "config", "study_area.yml")
    if (file.exists(yml)) cand else NULL
  },
  error = function(e) NULL
)
if (!is.null(infer_root_from_rmd)) {
  pr <- params$project_root
  is_default <- identical(pr, ".") || (is.character(pr) && length(pr) == 1L && pr %in% c(".", "./"))
  n_here <- length(Sys.glob(file.path(normalizePath(file.path(root_param, outd), winslash = "/", mustWork = FALSE), "maps", "*.png"))) +
    length(Sys.glob(file.path(normalizePath(file.path(root_param, outd), winslash = "/", mustWork = FALSE), "charts", "*.png")))
  n_inf <- length(Sys.glob(file.path(normalizePath(file.path(infer_root_from_rmd, outd), winslash = "/", mustWork = FALSE), "maps", "*.png"))) +
    length(Sys.glob(file.path(normalizePath(file.path(infer_root_from_rmd, outd), winslash = "/", mustWork = FALSE), "charts", "*.png")))
  if (is_default || n_inf > n_here) {
    root_param <- infer_root_from_rmd
  }
}
# Candidate explicit outputs dir from params (may be wrong if project_root was getwd() = docs/).
oroot_explicit <- NULL
if (!is.null(params$outputs_root) && nzchar(as.character(params$outputs_root))) {
  oroot_explicit <- normalizePath(as.character(params$outputs_root), winslash = "/", mustWork = FALSE)
}
# Resolve true repo root: "." under knit often points at docs/, so config/ and data/processed
# are missed. Prefer parent of this Rmd, then any candidate whose configured outputs folder has PNGs.
repo_from_rmd <- tryCatch({
  rd <- knitr::current_input(dir = TRUE)
  if (is.null(rd) || !nzchar(rd)) {
    ""
  } else {
    rd <- normalizePath(rd, winslash = "/", mustWork = FALSE)
    normalizePath(file.path(rd, ".."), winslash = "/", mustWork = FALSE)
  }
}, error = function(e) "")
candidates <- unique(c(
  repo_from_rmd,
  root_param,
  normalizePath(file.path(root_param, ".."), winslash = "/", mustWork = FALSE)
))
candidates <- candidates[vapply(candidates, function(r) {
  nzchar(r) && !is.na(r) && dir.exists(r)
}, logical(1))]
count_output_pngs <- function(r) {
  odir <- normalizePath(file.path(r, outd), winslash = "/", mustWork = FALSE)
  length(Sys.glob(file.path(odir, "maps", "*.png"))) +
    length(Sys.glob(file.path(odir, "charts", "*.png")))
}
root <- root_param
picked <- FALSE
for (r in candidates) {
  if (count_output_pngs(r) > 0L) {
    root <- r
    picked <- TRUE
    break
  }
}
if (!picked) {
  for (r in candidates) {
    if (dir.exists(normalizePath(file.path(r, outd), winslash = "/", mustWork = FALSE))) {
      root <- r
      break
    }
  }
}
# Phase 7 PNGs live under picked project root, but render() often passes outputs_root built from
# getwd() (e.g. .../docs/outputs) while root above is the real repo — then has_o() is always FALSE,
# figure chunks are skipped, and the HTML has empty Results subsections. Prefer the directory
# that actually contains maps/*.png + charts/*.png.
oroot_under_root <- normalizePath(file.path(root, outd), winslash = "/", mustWork = FALSE)
oroot_candidates <- unique(c(oroot_under_root, oroot_explicit))
oroot_candidates <- oroot_candidates[vapply(oroot_candidates, function(p) {
  nzchar(p) && !is.na(p)
}, logical(1))]
count_phase7_pngs <- function(d) {
  if (!length(d) || !nzchar(d) || is.na(d) || !dir.exists(d)) {
    return(0L)
  }
  length(Sys.glob(file.path(d, "maps", "*.png"))) +
    length(Sys.glob(file.path(d, "charts", "*.png")))
}
png_counts <- vapply(oroot_candidates, count_phase7_pngs, integer(1))
if (length(png_counts) && max(png_counts) > 0L) {
  oroot <- oroot_candidates[[which.max(png_counts)]]
} else {
  oroot <- oroot_under_root
}
png_n_maps <- length(Sys.glob(file.path(oroot, "maps", "*.png")))
png_n_charts <- length(Sys.glob(file.path(oroot, "charts", "*.png")))
if (knitr::is_html_output() && png_n_maps + png_n_charts == 0L) {
  message(
    "\n*** urban_sprawl_report: no PNG files in ", file.path(outd, "maps|charts"), ".\n",
    "    Project root: ", root, "\n",
    "    Outputs dir: ", oroot, "\n",
    "    This report embeds only those PNGs (from Phase 7). GeoTIFFs (.tif) and CSVs in outputs/\n",
    "    do not appear as figures until Phase 7 exports maps/charts.\n",
    "    Run: source(\"R/07_visualization.R\"); run_phase7(render_report = FALSE)\n",
    "    Then re-render this report. ***\n"
  )
}

# Paths for knitr::include_graphics() must be relative to knit_root_dir when set so rmarkdown can
# copy into *_files/ and pandoc --embed-resources inlines them (reliable HTML). Base64 asis_output
# was dropped by some Pandoc/markdown paths.
knit_root_dir_get <- function() {
  rd <- knitr::opts_knit$get("root.dir")
  if (!length(rd) || !nzchar(rd[[1L]])) {
    return("")
  }
  normalizePath(rd[[1L]], winslash = "/", mustWork = FALSE)
}
abs_to_knit_rel <- function(abs_paths) {
  abs_paths <- normalizePath(as.character(abs_paths), winslash = "/", mustWork = FALSE)
  kr <- knit_root_dir_get()
  if (!nzchar(kr)) {
    out <- gsub("\\\\", "/", abs_paths)
    return(out)
  }
  kr <- normalizePath(kr, winslash = "/", mustWork = FALSE)
  vapply(abs_paths, function(p) {
    p <- normalizePath(p, winslash = "/", mustWork = FALSE)
    pre <- paste0(kr, "/")
    nc <- nchar(pre)
    rel <- if (nchar(p) >= nc && identical(tolower(substr(p, 1L, nc)), tolower(pre))) {
      substr(p, nchar(kr) + 2L, nchar(p))
    } else {
      p
    }
    gsub("\\\\", "/", rel)
  }, character(1), USE.NAMES = FALSE)
}
include_project_graphics <- function(paths) {
  paths <- as.character(paths)
  paths <- paths[nzchar(paths)]
  if (!length(paths)) {
    return(invisible(NULL))
  }
  knit_rd <- knitr::opts_knit$get("root.dir")
  resolve_graphic_path <- function(p) {
    p <- as.character(p)[1L]
    if (!length(knit_rd) || !nzchar(knit_rd[[1L]])) {
      return(normalizePath(p, winslash = "/", mustWork = FALSE))
    }
    if (grepl("^/|^[A-Za-z]:[\\\\/]", p)) {
      return(normalizePath(p, winslash = "/", mustWork = FALSE))
    }
    normalizePath(file.path(knit_rd[[1L]], p), winslash = "/", mustWork = FALSE)
  }
  paths_abs <- vapply(paths, resolve_graphic_path, character(1), USE.NAMES = FALSE)
  miss <- !file.exists(paths_abs)
  if (any(miss)) {
    stop("Missing figure file(s): ", paste(paths[miss], collapse = ", "), call. = FALSE)
  }
  paths_use <- paths_abs
  if (knitr::is_latex_output() && identical(.Platform$OS.type, "windows")) {
    paths_use <- vapply(paths_abs, function(p) {
      sp <- tryCatch(
        normalizePath(utils::shortPathName(p), winslash = "/", mustWork = FALSE),
        error = function(e) ""
      )
      if (nzchar(sp) && file.exists(sp)) sp else p
    }, character(1), USE.NAMES = FALSE)
  }
  rels <- abs_to_knit_rel(paths_use)
  knitr::include_graphics(rels)
  invisible(NULL)
}

# Absolute path to a file under the project (for existence checks).
fig_abs <- function(...) {
  normalizePath(file.path(root, ...), winslash = "/", mustWork = FALSE)
}
has <- function(...) file.exists(fig_abs(...))
# Single relpath for chunk option eval=... — knitr splits options on commas, so
# eval=has("outputs","maps","x.png") is parsed as multiple options and breaks.
has_rel <- function(relpath) {
  file.exists(normalizePath(file.path(root, relpath), winslash = "/", mustWork = FALSE))
}
# Figures and many CSVs live under Phase 7 outputs (oroot), not necessarily root/outputs/.
has_o <- function(path_under_outputs) {
  p <- tryCatch(
    normalizePath(file.path(oroot, path_under_outputs), winslash = "/", mustWork = FALSE),
    error = function(e) ""
  )
  if (!nzchar(p) || (length(p) != 1L) || is.na(p)) {
    return(FALSE)
  }
  isTRUE(file.exists(p))
}
fig_o <- function(path_under_outputs) {
  p <- normalizePath(file.path(oroot, path_under_outputs), winslash = "/", mustWork = FALSE)
  if (!file.exists(p)) return(p)
  if (knitr::is_latex_output() && identical(.Platform$OS.type, "windows")) {
    sp <- tryCatch(
      normalizePath(utils::shortPathName(p), winslash = "/", mustWork = FALSE),
      error = function(e) ""
    )
    if (nzchar(sp) && file.exists(sp)) return(sp)
  }
  p
}

# HTML + PDF: use include_graphics(); HTML paths are relative to knit_root_dir for self-contained.
ig_outputs <- function(path_under_outputs) {
  abs_p <- normalizePath(file.path(oroot, path_under_outputs), winslash = "/", mustWork = FALSE)
  if (!file.exists(abs_p)) {
    knitr::asis_output(paste0("\n\n*Missing figure: `", file.path(outd, path_under_outputs), "`.*\n\n"))
    return(invisible(NULL))
  }
  if (knitr::is_html_output()) {
    knitr::include_graphics(abs_to_knit_rel(abs_p))
  } else {
    knitr::include_graphics(fig_o(path_under_outputs))
  }
  invisible(NULL)
}

# Paths passed to knitr::include_graphics() must be LaTeX-safe. Spaces in Windows
# paths (e.g. C:/Urban Sprawl/...) break \includegraphics unless quoted; knitr
# may not always quote. Use 8.3 short paths for PDF on Windows.
fig <- function(...) {
  p <- fig_abs(...)
  if (!file.exists(p)) return(p)
  if (knitr::is_latex_output() && identical(.Platform$OS.type, "windows")) {
    sp <- tryCatch(
      normalizePath(utils::shortPathName(p), winslash = "/", mustWork = FALSE),
      error = function(e) ""
    )
    if (nzchar(sp) && file.exists(sp)) return(sp)
  }
  p
}

path_proc <- function(...) {
  normalizePath(file.path(root, "data", "processed", ...), winslash = "/", mustWork = FALSE)
}
rp_ch <- file.path(root, "R", "report_knit_charts.R")
if (!file.exists(rp_ch)) {
  rp_alt <- file.path(root_param, "R", "report_knit_charts.R")
  if (file.exists(rp_alt)) rp_ch <- rp_alt
}
if (file.exists(rp_ch)) {
  sys.source(rp_ch, envir = knitr::knit_global())
}
has_knit_fn <- function(nm) {
  exists(nm, envir = knitr::knit_global(), mode = "function", inherits = FALSE)
}

if (knitr::is_latex_output()) {
  knitr::opts_chunk$set(out.width = "0.92\\linewidth")
} else if (knitr::is_html_output()) {
  knitr::opts_chunk$set(out.width = "100%", out.extra = 'class="report-fig-img"')
} else {
  knitr::opts_chunk$set(out.width = "92%")
}
```

```{css, echo=FALSE, eval=knitr::is_html_output()}
/*
 * IEEE / MDPI–inspired article layout (HTML): serif body, sans headings,
 * MDPI-style green + IEEE blue accents, structured abstract block, TOC on top.
 */
/* Crimson Pro + IBM Plex Sans (journal). Falls back to Times/system if fonts.googleapis.com is blocked. */
@import url("https://fonts.googleapis.com/css2?family=Crimson+Pro:ital,wght@0,400;0,600;0,700;1,400&family=IBM+Plex+Sans:ital,wght@0,400;0,500;0,600;0,700;1,400&display=swap");

html { scroll-padding-top: 4.5rem; }

/* Journal masthead strip (MDPI green + IEEE blue) */
body::before {
  content: "";
  display: block;
  height: 5px;
  background: linear-gradient(90deg, #1a6f54 0%, #1a6f54 42%, #00629b 42%, #00629b 100%);
}

body {
  font-family: "Crimson Pro", "Times New Roman", Times, "Liberation Serif", Georgia, serif;
  font-size: 17px;
  line-height: 1.65;
  margin: 0;
  padding: 0 0 3rem;
  color: #222;
  -webkit-font-smoothing: antialiased;
}

a, a:focus { color: #00629b; text-decoration: none; }
a:hover { color: #004870; text-decoration: underline; }

code, kbd, pre, samp {
  font-family: Consolas, "Liberation Mono", "Courier New", monospace;
  font-size: 0.88em;
}
div.toc-content :not(pre) > code {
  background: #eef4f2;
  padding: 0.1em 0.35em;
  border-radius: 2px;
  color: #1a1a1a;
}

/* Page shell — single column; avoid Bootstrap 3 float grid splitting TOC vs body. */
div.main-container.container-fluid {
  max-width: 52rem;
  margin-left: auto;
  margin-right: auto;
  padding: 1rem clamp(0.75rem, 3vw, 1.5rem) 3rem;
  background: #eceff1;
  box-sizing: border-box;
}
div.main-container.container-fluid > .row,
div.main-container.container-fluid .row {
  display: block !important;
  margin-left: 0 !important;
  margin-right: 0 !important;
}
div.main-container.container-fluid .row > [class*="col-"] {
  float: none !important;
  width: 100% !important;
  max-width: 100% !important;
  clear: both !important;
  padding-left: 0 !important;
  padding-right: 0 !important;
}

/* TOC on top (toc_float: false): full-width panel, then one article column */
#TOC {
  float: none !important;
  width: 100% !important;
  max-width: 100% !important;
  position: relative !important;
  margin: 0 0 1.5rem 0 !important;
  font-family: "IBM Plex Sans", system-ui, sans-serif !important;
  font-size: 0.85rem;
  line-height: 1.45;
  background: #f6faf8 !important;
  border: 1px solid #b8dcc8;
  border-radius: 4px;
  padding: 0.65rem 1rem 0.85rem;
  box-shadow: 0 1px 2px rgba(0, 0, 0, 0.06);
  box-sizing: border-box;
}
#TOC ul {
  margin: 0.35rem 0 0 0;
  padding-left: 1.1rem;
}
#TOC a {
  color: #00629b;
}

/* Main body: single column, readable measure (IEEE/MDPI–like flow) */
div.toc-content {
  float: none !important;
  width: 100% !important;
  max-width: 100% !important;
  margin: 0 !important;
  padding: 2rem 2.25rem 2.5rem !important;
  box-sizing: border-box;
  background: #fff;
  border-radius: 4px;
  box-shadow: 0 2px 12px rgba(0, 0, 0, 0.08);
  border: 1px solid #dde3ea;
  clear: both;
}

/* Pandoc sections must stack (never sit side-by-side) */
div.toc-content .section,
div.toc-content .section.level1,
div.toc-content .section.level2,
div.toc-content .section.level3 {
  float: none !important;
  clear: both !important;
  display: block !important;
  width: 100% !important;
  max-width: 100% !important;
  box-sizing: border-box;
  position: relative;
}
div.toc-content .section.level1 {
  margin-bottom: 2.5rem;
}

/* Title block — centered article header */
#header {
  text-align: center;
  max-width: 100%;
  margin-left: auto;
  margin-right: auto;
  padding: 1.25rem 0.75rem 1.5rem;
  margin-bottom: 1.5rem;
  border-bottom: 3px double #2c3e50;
  box-sizing: border-box;
  clear: both;
  float: none !important;
}
#header::before {
  content: "Research article";
  display: block;
  font-family: "IBM Plex Sans", system-ui, sans-serif;
  font-size: 0.68rem;
  font-weight: 600;
  text-transform: uppercase;
  letter-spacing: 0.22em;
  color: #1a6f54;
  margin-bottom: 1rem;
}
h1.title {
  font-family: "IBM Plex Sans", system-ui, sans-serif;
  font-weight: 700;
  font-size: clamp(1.3rem, 2.4vw, 1.8rem);
  color: #111;
  line-height: 1.28;
  letter-spacing: -0.02em;
  max-width: none;
  margin-top: 0;
}
.subtitle {
  font-family: "IBM Plex Sans", system-ui, sans-serif;
  color: #424242;
  font-size: 0.98rem;
  font-weight: 500;
  margin-top: 0.5rem;
  line-height: 1.45;
}
#header .author,
#header .date {
  font-family: "IBM Plex Sans", system-ui, sans-serif;
  font-size: 0.84rem;
  color: #555;
  margin-top: 0.65rem;
}

.level1 { margin-top: 2.25rem; }

/* Section titles: sans, strong rules (IEEE-style clarity) */
div.toc-content h1 {
  font-family: "IBM Plex Sans", system-ui, sans-serif;
  font-size: 1.32rem;
  font-weight: 700;
  color: #111;
  border-bottom: 2px solid #222;
  padding-bottom: 0.35rem;
  margin-top: 2.5rem;
}
div.toc-content h2 {
  font-family: "IBM Plex Sans", system-ui, sans-serif;
  font-size: 1.12rem;
  font-weight: 600;
  color: #1a1a1a;
  margin-top: 2rem;
  border-bottom: 1px solid #ccc;
  padding-bottom: 0.25rem;
}
div.toc-content h3 {
  font-family: "IBM Plex Sans", system-ui, sans-serif;
  font-size: 1.02rem;
  font-weight: 600;
  color: #333;
  margin-top: 1.5rem;
}

/* Abstract + front matter (MDPI-style panel; explicit fenced div in .Rmd) */
div.toc-content .journal-abstract-section {
  background: linear-gradient(180deg, #f5fbf8 0%, #fafcfb 100%);
  border: 1px solid #b8dcc8;
  border-left: 5px solid #1a6f54;
  padding: 1.1rem 1.3rem 1.25rem;
  margin: 0 0 2.25rem;
  border-radius: 0 3px 3px 0;
  box-shadow: 0 1px 4px rgba(26, 111, 84, 0.08);
}
div.toc-content .journal-abstract-section h1 {
  font-size: 1.15rem;
  color: #1a6f54;
  border-bottom: none;
  margin-top: 0;
  padding-bottom: 0.25rem;
}
div.toc-content .journal-abstract-section h1 .header-section-number {
  color: #1a6f54;
}

/* Keywords line (MDPI-style) */
.journal-keywords {
  font-family: "IBM Plex Sans", system-ui, sans-serif;
  font-size: 0.86rem;
  margin-top: 1.1rem;
  padding-top: 1rem;
  border-top: 1px solid #a8d4bc;
  color: #333;
  line-height: 1.55;
}
.journal-keywords strong { color: #1a6f54; font-weight: 700; }

blockquote {
  border-left: 4px solid #00629b;
  padding: 0.5rem 1rem;
  margin: 1.1rem 0;
  color: #37474f;
  background: #f4f8fb;
  font-size: 0.96em;
}

/*
 * Figures — IEEE / MDPI–style: single column width, symmetric margins, ruled
 * panel, caption band matching figure width (no narrower caption than graphic).
 */
div.toc-content div.figure {
  display: flex;
  flex-direction: column;
  align-items: stretch;
  width: 100%;
  max-width: 100%;
  margin: 2rem auto;
  padding: 0.85rem 1rem 0.65rem;
  box-sizing: border-box;
  text-align: center;
  background: #fafbfc;
  border: 1px solid #cfd6de;
  border-top: 2px solid #1a6f54;
  box-shadow: 0 1px 0 rgba(0, 0, 0, 0.04);
}
div.toc-content div.figure img,
div.toc-content img.report-fig-img {
  width: 100% !important;
  max-width: 100% !important;
  height: auto !important;
  margin: 0 auto;
  display: block;
  border: 1px solid #b0b8c2;
  box-sizing: border-box;
}
div.toc-content div.figure img {
  background: #fff;
}
div.toc-content p img.report-fig-img {
  width: 100% !important;
  max-width: 100% !important;
}
/* Multi-panel change maps: equal cell width, symmetric gutters (MDPI-style grid) */
div.toc-content .journal-figure-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(min(100%, 16.5rem), 1fr));
  gap: 1rem 1.15rem;
  width: 100%;
  max-width: 100%;
  margin: 2rem auto;
  padding: 1rem 1rem 0.85rem;
  box-sizing: border-box;
  background: #fafbfc;
  border: 1px solid #cfd6de;
  border-top: 2px solid #00629b;
  justify-items: stretch;
  align-items: start;
}
div.toc-content .journal-figure-grid > p {
  margin: 0;
  display: block;
  width: 100%;
  text-align: center;
}
div.toc-content .journal-figure-grid img {
  width: 100% !important;
  max-width: 100% !important;
  height: auto !important;
  display: block;
  margin: 0 auto;
  border: 1px solid #b0b8c2;
  box-sizing: border-box;
  background: #fff;
}
/* If knitr wraps each map in .figure, strip outer panel so grid stays symmetric */
div.toc-content .journal-figure-grid .figure {
  background: transparent !important;
  border: none !important;
  box-shadow: none !important;
  padding: 0 !important;
  margin: 0 !important;
  border-top: none !important;
  width: 100%;
  max-width: 100%;
}
div.toc-content .journal-figure-grid .figure img {
  border: 1px solid #b0b8c2;
}
div.toc-content .journal-figure-grid .figure p.caption {
  border-top: none;
  font-size: 0.72rem;
  margin-top: 0.35rem;
  padding-top: 0;
}
/* Captions: full width of figure panel; justified with centered last line (common in journals) */
div.toc-content p.caption,
div.toc-content .figure p.caption,
div.toc-content .figure .caption,
div.toc-content figcaption {
  font-family: "IBM Plex Sans", system-ui, sans-serif;
  font-size: 0.78rem;
  font-weight: 400;
  color: #333;
  text-align: justify;
  text-align-last: center;
  hyphens: auto;
  -webkit-hyphens: auto;
  width: 100%;
  max-width: 100%;
  margin: 0.65rem 0 0;
  padding: 0.55rem 0.15rem 0;
  line-height: 1.48;
  border-top: 1px solid #dde3ea;
  box-sizing: border-box;
}
div.toc-content p.caption strong:first-child,
div.toc-content .figure p.caption strong:first-child {
  font-weight: 700;
  color: #111;
}

/* Tables — real table layout (prevents columns splitting across the page) */
div.toc-content .table-responsive,
div.toc-content .section {
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
div.toc-content table {
  display: table !important;
  width: 100%;
  max-width: 100%;
  table-layout: auto;
  font-size: 0.92rem;
  border-collapse: collapse;
  margin: 1.25rem 0;
  font-variant-numeric: tabular-nums;
  border: 1px solid #cfd6de;
  background: #fff;
}
div.toc-content table th {
  font-family: "IBM Plex Sans", system-ui, sans-serif;
  font-weight: 600;
  border: 1px solid #cfd6de;
  border-bottom: 2px solid #222;
  padding: 0.5rem 0.7rem;
  text-align: left;
  background: #f0f4f8;
  white-space: nowrap;
}
div.toc-content table td {
  border: 1px solid #e2e8f0;
  padding: 0.5rem 0.7rem;
  vertical-align: top;
}
div.toc-content table caption {
  caption-side: top;
  text-align: left;
  font-family: "IBM Plex Sans", system-ui, sans-serif;
  font-weight: 600;
  font-size: 0.9rem;
  padding: 0 0 0.5rem 0;
  color: #1a1a1a;
}

div.toc-content .html-widget {
  max-width: 100% !important;
  width: 100% !important;
  margin-left: auto !important;
  margin-right: auto !important;
  box-sizing: border-box;
}

/* paged / DT tables (if any) must not escape the article column */
div.toc-content .pagedtable,
div.toc-content .pagedtable-wrapper,
div.toc-content .datatables,
div.toc-content div.dataTables_wrapper {
  max-width: 100% !important;
  overflow-x: auto !important;
  box-sizing: border-box;
}

@media (max-width: 767px) {
  div.main-container.container-fluid {
    padding: 0.5rem;
  }
  div.toc-content {
    padding: 1.25rem 1rem 2rem !important;
  }
  #header {
    padding-left: 0.5rem;
    padding-right: 0.5rem;
  }
}
```

:::: {.journal-abstract-section}

# Abstract

This document is the main **research-style written deliverable** for the Urban Sprawl pipeline. It summarizes methods, **actual outputs** embedded from `outputs/` (maps, charts, tables), and an **honest** statement of limitations and **research gaps** (what we did *not* establish—e.g. causality, independent validation of the CNN, full-resolution drivers). The **PDF** version (`urban_sprawl_report.pdf`) is intended for submission, archiving, or thesis appendices; **HTML** is for quick browsing. A technical runbook is `docs/PROJECT_IMPLEMENTATION.txt`; an interactive map is `outputs/urban_sprawl_dashboard.html`.

This report documents an end-to-end **reproducible workflow** for mapping multi-temporal land cover, detecting new urban expansion with a deep-learning change model, quantifying sprawl-style patterns (Shannon entropy, UEII, landscape metrics, hotspots), and exploring **associations** between binary urban growth and coarse-scale accessibility / environmental predictors. Optical imagery is sourced from **Microsoft Planetary Computer** (Sentinel-2 and Landsat) and processed in **R** with **terra**; classification uses **randomForest** and **e1071** (SVM); change detection uses a **torch** Siamese convolutional network; sprawl diagnostics use **sf**-based zonal geometry and **landscapemetrics**; driver exploration uses logistic regression and correlation on **stratified samples** (not full-census inference).

**Study area:** `r params$study_name`

::: {.journal-keywords}
**Keywords:** urban sprawl; land-cover change; deep learning; Shannon entropy; remote sensing; R; reproducible analysis
:::

::::

<hr />

# Introduction

Urban expansion alters ecosystems, hydrology, and travel patterns. Policy and planning benefit from **repeatable** maps and metrics that separate **where** growth occurred from **how concentrated** it is across distance from the core. This project implements a seven-phase pipeline from imagery to maps, charts, and a web dashboard.

## Research context

Regional studies often combine optical time series, pixel- or object-based classification, and landscape pattern indices. **Change detection** ranges from post-classification comparison to dedicated neural architectures (e.g. Siamese encoders) that learn from bi-temporal feature stacks. **Sprawl** is frequently operationalized via entropy of the spatial distribution of urban area across zones, or via patch/edge metrics from landscape ecology.

## Honest scope of this work

What this pipeline **does** deliver:

- Consistent AOI-wide **LULC** layers for configured years, with documented class codes (1–4).
- **CNN-derived** binary “new urban” layers for each configured pre/post pair (subject to training data, spatial split, and thresholding).
- **Zonal** Shannon entropy, UEII, landscape metrics for class 1 (urban), and hotspot polygons under explicit rules.
- **Exploratory** driver analysis (logistic regression on samples; optional spatial lag model; GWR off by default) on a **coarsened** grid to remain tractable.

What it **does not** claim:

- **Causal** effects of roads or distance to CBD on growth (models are associational; endogeneity is not addressed).
- **Validated** CNN transfer to cities or years outside the training design without re-training.
- **Full-resolution** driver surfaces in Phase 6: distances and stacks are computed on an aggregated grid (`phase6.coarsen_fact` / `working_resolution_m`) by design.
- **Complete** road network representation: OSM coverage varies; failed downloads fall back to placeholder distance fields.

## Research gaps (explicit)

1. **Reference data:** Automated training from WorldCover (or similar) is convenient but may **misalign** with local definitions of “urban”; independent reference labels would strengthen accuracy claims.
2. **Change validation:** CNN metrics on **patch splits** do not replace **pixel- or object-level** validation against independent change reference.
3. **Sprawl theory:** Entropy and UEII are **indicators**, not a single definition of “sprawl”; multi-scale and temporal stationarity assumptions deserve sensitivity analysis.
4. **Drivers:** Sampled logistic models omit unobserved drivers (policy, land market, infrastructure plans). **Spatial autocorrelation** of residuals may remain after a lag model; GWR is optional and heavy.
5. **Generalization:** Results are **case-specific** to the configured bbox, seasons, and cloud masks; external validity requires re-running or adapting the pipeline.

## What was implemented vs. what remains open (summary)

| Topic | Implemented in this repository | Still open / “gap” |
|-------|-------------------------------|----------------------|
| Imagery → features | PC STAC, signing, masking, indices, normalization | Season/cloud sensitivity; no automatic full-year S2 fallback |
| LULC | RF + SVM, optional auto-training from WorldCover | Independent reference validation; local urban definition |
| Change | Siamese CNN, spatial block splits, multi-transition | Patch-level metrics are not full-map accuracy; CPU runtime |
| Sprawl metrics | Rings, Shannon, UEII, landscapemetrics, hotspots | Alternative sprawl definitions; multi-scale sensitivity |
| Drivers | Coarsened grid, OSM cache, GLM / Moran / lagsarlm | Causality; unobserved confounders; optional GWR cost |
| Reporting | Static maps/charts, Leaflet, **PDF + HTML** report | Dashboard is exploratory, not a peer-review figure |

<hr />

# Objectives

1. Build analysis-ready multi-temporal **feature** and **LULC** stacks for the study AOI.  
2. Produce **binary new-urban** maps for each configured epoch pair via the Siamese CNN.  
3. Summarize **sprawl-relevant** zonal and landscape metrics and identify **hotspot** zones under stated rules.  
4. Document **associations** between growth and coarse-scale predictors, with clear limits on interpretation.

<hr />

# Study area and configuration

All geographic and processing choices live in **`config/study_area.yml`**: WGS84 bounding box, projected CRS (e.g. UTM), Planetary Computer collections and years, CNN transition list, sprawl centre coordinates, Phase 6/7 options. **Reproducibility** requires versioning or archiving that file alongside outputs.

```{r flowchart, eval=FALSE, fig.cap="Figure 1. Seven-phase workflow (repository methodology figure)."}
# Omitted by default (optional asset). To include: add docs/methodology-flowchart.png and set eval=TRUE.
include_project_graphics(fig("docs", "methodology-flowchart.png"))
```

<hr />

# Materials and methods (summary)

| Phase | Role | Main tools |
|------:|------|------------|
| 1 | Imagery acquisition & stacking | rstac, PC signing, terra |
| 2 | Cloud/shadow masking, indices, normalization | terra |
| 3 | LULC (RF + SVM), rules, area stats | randomForest, e1071, caret |
| 4 | Siamese CNN change detection | torch |
| 5 | Rings, Shannon H, UEII, landscape metrics, hotspots | sf, landscapemetrics |
| 6 | Driver rasters, stratified sample, GLM / spatial models | terra, osmdata, spdep, spatialreg |
| 7 | Static maps, charts, Leaflet, this report | tmap, ggplot2, leaflet, rmarkdown |

**Phase 6 note:** Predictor stacks are built on a **coarsened** LULC grid to control compute; interpretations refer to that scale, not necessarily native 10 m pixels.

<hr />

# Results

**Figures** normally come from **`outputs/maps/*.png`** and **`outputs/charts/*.png`**, written by **`run_phase7()`** (tmap/ggplot). If those PNGs are missing, this report **builds the same style of charts from CSVs at knit time** (caption says “knit time”) so Results stay readable. **Maps** (LULC panel, change, hotspots) still need Phase 7 exports and underlying GeoTIFFs. **Tables** use `data/processed/*.csv` and **`outputs/*.csv`**. Always run **`setwd()` to the project root** (or open **`UrbanSprawl.Rproj`**) before **`run_phase7()`** so files land in the real `outputs/` folder.

```{r results-banner, echo=FALSE, results='asis'}
if (png_n_maps + png_n_charts == 0L) {
  knitr::asis_output(paste0(
    "\n\n> **Note:** The folders `", file.path(outd, "maps"), "` and `", file.path(outd, "charts"),
    "` contain no PNG files yet. Chart-style figures below are built from CSVs at knit time when possible. ",
    "From the **project root** run `source(\"R/07_visualization.R\"); run_phase7()` to export high-DPI maps and charts.\n\n"
  ))
}
```

```{r outputs-inventory, echo=FALSE}
inv_dir <- oroot
if (!dir.exists(inv_dir)) {
  knitr::asis_output(paste0(
    "\n\n*Outputs inventory skipped: directory does not exist: `", inv_dir, "`*\n\n"
  ))
} else {
  fl <- list.files(inv_dir, pattern = "\\.(png|tif|csv)$", full.names = FALSE, recursive = TRUE, ignore.case = TRUE)
  fl <- sort(unique(fl))
  fl <- fl[!grepl("^urban_sprawl_dashboard_files/", fl) & !grepl("^urban_sprawl_report_files/", fl)]
  if (length(fl)) {
    nmax <- 60L
    cap <- paste0(
      "Inventory of .png / .tif / .csv under outputs/ (showing first ", min(nmax, length(fl)), " of ", length(fl),
      " paths). Only PNGs in maps/ and charts/ are embedded as figures above."
    )
    knitr::kable(
      data.frame(`File (under outputs/)` = head(fl, nmax), check.names = FALSE),
      caption = cap
    )
    if (length(fl) > nmax) {
      knitr::asis_output(paste0("\n\n*…and ", length(fl) - nmax, " more paths under `outputs/`.*\n\n"))
    }
  } else {
    knitr::asis_output(paste0(
      "\n\n*No .png / .tif / .csv found under `", inv_dir, "` (recursive). Check Phase 7 and `oroot`.*\n\n"
    ))
  }
}
```

## Land cover — multi-year panel

```{r lulc-fig, echo=FALSE}
if (has_o("maps/lulc_panel_map.png")) {
  knitr::asis_output("**Figure 2.** Classified LULC panel (Phase 3).\n\n")
  ig_outputs("maps/lulc_panel_map.png")
} else {
  knitr::asis_output(paste0("\n\n*Figure 2 omitted: run Phase 3 and Phase 7 to create `", file.path(outd, "maps/lulc_panel_map.png"), "`.*\n\n"))
}
```

## Urban area trajectory

```{r growth-png, echo=FALSE}
if (has_o("charts/urban_growth_trajectory.png")) {
  knitr::asis_output("**Figure 3.** Urban class area (km²) by year (Phase 3; chart PNG from Phase 7).\n\n")
  ig_outputs("charts/urban_growth_trajectory.png")
}
```

```{r growth-ggplot, fig.width=10, fig.height=6, dev="png", dpi=144, echo=FALSE, fig.cap="Figure 3. Urban class area (km²) by year from processed class-area statistics (Phase 3; PNG from Phase 7 if present, otherwise generated at knit time)."}
if (has_o("charts/urban_growth_trajectory.png")) {
  invisible(NULL)
} else if (has_knit_fn("knit_urban_growth_plot") && file.exists(path_proc("urban_area_stats.csv"))) {
  gp <- knit_urban_growth_plot(root, params$study_name)
  if (!is.null(gp)) {
    print(gp)
  } else {
    knitr::asis_output("*Figure 3: could not build chart from processed urban-area statistics CSV.*")
  }
} else {
  knitr::asis_output("*Figure 3 omitted: missing CSV, PNG, or `R/report_knit_charts.R` (for knit-time plots).*")
}
```

## Shannon entropy (sprawl indicator)

```{r entropy-png, echo=FALSE}
if (has_o("charts/shannon_entropy_chart.png")) {
  knitr::asis_output("**Figure 4.** Relative Shannon entropy across buffer zones (Phase 5).\n\n")
  ig_outputs("charts/shannon_entropy_chart.png")
}
```

```{r entropy-ggplot, fig.width=10, fig.height=6, dev="png", dpi=144, echo=FALSE, fig.cap="Figure 4. Relative Shannon entropy across buffer zones (Phase 5)."}
if (has_o("charts/shannon_entropy_chart.png")) {
  invisible(NULL)
} else if (has_knit_fn("knit_entropy_plot") && file.exists(path_proc("shannon_entropy_results.csv"))) {
  ep <- knit_entropy_plot(root)
  if (!is.null(ep)) {
    print(ep)
  } else {
    knitr::asis_output("*Figure 4: could not build from `shannon_entropy_results.csv`.*")
  }
} else {
  knitr::asis_output("*Figure 4 omitted: missing Shannon entropy CSV, PNG, or knit helpers.*")
}
```

## CNN change maps (new urban)

```{r change-maps, echo=FALSE}
fps <- character()
cfg_rmd <- tryCatch(
  yaml::read_yaml(fig_abs("config", "study_area.yml")),
  error = function(e) NULL
)
trans <- if (is.null(cfg_rmd) || is.null(cfg_rmd$cnn)) list() else cfg_rmd$cnn$transitions
if (is.null(trans)) trans <- list()
for (t in trans) {
  sub <- sprintf("change_map_%s_%s.png", t$pre, t$post)
  relm <- paste("maps", sub, sep = "/")
  if (has_o(relm)) {
    fps <- c(fps, normalizePath(file.path(oroot, relm), winslash = "/", mustWork = FALSE))
  }
}
if (length(fps)) {
  knitr::asis_output("**Change maps (new urban).** CNN-derived transitions (Phase 4/7).\n\n")
  include_project_graphics(fps)
} else {
  knitr::asis_output("*No change-map PNGs found. Run Phase 4 and Phase 7 (`save_change_maps`).*")
}
```

## Sprawl hotspots (map)

```{r hotspots-fig, echo=FALSE}
if (has_o("maps/sprawl_hotspot_map.png")) {
  knitr::asis_output("**Figure 5.** Latest-year LULC with Phase 5 hotspot zones (if produced).\n\n")
  ig_outputs("maps/sprawl_hotspot_map.png")
} else {
  knitr::asis_output("\n\n*Figure 5 omitted: run Phase 5 and Phase 7 to create `outputs/maps/sprawl_hotspot_map.png`.*\n\n")
}
```

## Spatial drivers and correlations

```{r drivers-png, echo=FALSE}
if (has_o("charts/driver_importance_chart.png")) {
  knitr::asis_output("**Figure 6.** Logistic regression coefficient magnitudes (Phase 6 sample; associational).\n\n")
  ig_outputs("charts/driver_importance_chart.png")
}
```

```{r drivers-ggplot, fig.width=10, fig.height=7, fig.cap="Figure 6. Logistic regression coefficient magnitudes (Phase 6 sample; associational).", dev="png", dpi=144, echo=FALSE}
if (has_o("charts/driver_importance_chart.png")) {
  invisible(NULL)
} else if (has_knit_fn("knit_drivers_plot") && file.exists(file.path(oroot, "logistic_regression_results.csv"))) {
  dp <- knit_drivers_plot(oroot)
  if (!is.null(dp)) {
    print(dp)
  } else {
    knitr::asis_output("*Figure 6: could not plot driver results.*")
  }
} else {
  knitr::asis_output("*Figure 6 omitted: missing `logistic_regression_results.csv` and chart PNG.*")
}
```

```{r corr-png, echo=FALSE}
if (has_o("charts/correlation_heatmap.png")) {
  knitr::asis_output("**Figure 7.** Pearson correlation among sampled response and predictors.\n\n")
  ig_outputs("charts/correlation_heatmap.png")
}
```

```{r corr-ggplot, fig.width=9, fig.height=8, fig.cap="Figure 7. Pearson correlation among sampled response and predictors.", dev="png", dpi=144, echo=FALSE}
if (has_o("charts/correlation_heatmap.png")) {
  invisible(NULL)
} else if (has_knit_fn("knit_corr_plot") && requireNamespace("ggcorrplot", quietly = TRUE) && file.exists(file.path(oroot, "correlation_matrix.csv"))) {
  cp <- knit_corr_plot(oroot)
  if (!is.null(cp)) {
    print(cp)
  } else {
    knitr::asis_output("*Figure 7: could not build correlation plot.*")
  }
} else {
  knitr::asis_output("*Figure 7 omitted: install `ggcorrplot` and ensure `correlation_matrix.csv` exists, or run Phase 7 for PNG.*")
}
```

## Landscape metrics

```{r lsm-png, echo=FALSE}
if (has_o("charts/landscape_metrics_chart.png")) {
  knitr::asis_output("**Figure 8.** Patch density, LPI, edge density (urban class; Phase 5).\n\n")
  ig_outputs("charts/landscape_metrics_chart.png")
}
```

```{r lsm-ggplot, fig.width=11, fig.height=4.5, fig.cap="Figure 8. Patch density, LPI, edge density (urban class; Phase 5).", dev="png", dpi=144, echo=FALSE}
if (has_o("charts/landscape_metrics_chart.png")) {
  invisible(NULL)
} else if (has_knit_fn("knit_landscape_plot") && file.exists(path_proc("landscape_metrics.csv"))) {
  lp <- knit_landscape_plot(root)
  if (!is.null(lp)) {
    print(lp)
  } else {
    knitr::asis_output("*Figure 8: could not plot landscape metrics.*")
  }
} else {
  knitr::asis_output("*Figure 8 omitted: missing `landscape_metrics.csv` and chart PNG.*")
}
```

<hr />

# Tables (key CSV outputs)

```{r tables-urban}
p_urban <- file.path(root, "data", "processed", "urban_area_stats.csv")
if (file.exists(p_urban)) {
  u <- utils::read.csv(p_urban, check.names = FALSE)
  knitr::kable(u, caption = "Table 1. Class areas (km²) by year (Phase 3).")
}
```

```{r tables-entropy}
p_ent <- file.path(root, "data", "processed", "shannon_entropy_results.csv")
if (file.exists(p_ent)) {
  e <- utils::read.csv(p_ent, check.names = FALSE)
  knitr::kable(head(e, 20L), caption = "Table 2. Shannon entropy results (first 20 rows).")
}
```

```{r tables-cnn}
p_cnn <- file.path(oroot, "cnn_accuracy_report.csv")
if (file.exists(p_cnn)) {
  cnn <- utils::read.csv(p_cnn, check.names = FALSE)
  knitr::kable(cnn, caption = "Table 3. CNN change-detection metrics by transition (patch-level; Phase 4).")
}
```

```{r tables-glm}
p_glm <- file.path(oroot, "logistic_regression_results.csv")
if (file.exists(p_glm)) {
  g <- utils::read.csv(p_glm, check.names = FALSE)
  knitr::kable(g, caption = "Table 4. Logistic regression (Phase 6); odds ratios and Wald statistics.")
}
```

```{r tables-corr}
p_corr <- file.path(oroot, "correlation_matrix.csv")
if (file.exists(p_corr)) {
  cm <- utils::read.csv(p_corr, check.names = FALSE)
  knitr::kable(
    head(cm, min(25L, nrow(cm))),
    caption = "Table 5. Pearson correlation matrix among sampled variables (Phase 6; first rows)."
  )
  if (nrow(cm) > 25L) {
    knitr::asis_output(paste0("\n\n*…", nrow(cm) - 25L, " more variable(s) in `correlation_matrix.csv`.*\n\n"))
  }
}
```

<hr />

# Discussion

Rising **urban area** in Table 1 and Figure 3 supports a narrative of expansion if consistent with visual change maps. **Relative entropy** (Figure 4) addresses *dispersion* of urban mass across rings: increases may suggest more polycentric or spread-out growth *given the ring geometry*, but zone design (width, CBD choice) strongly shapes the metric. **UEII** and hotspots identify zones where expansion intensity exceeds a stated rule—not universal “bad sprawl” labels.

Phase 6 coefficients (Figure 6–7) summarize **marginal associations** on a stratified sample after coarsening. Distance to roads or centre may correlate with growth for structural reasons that are not policy levers; readers should avoid causal language without a dedicated design.

<hr />

# Limitations and future work

- **Data:** Cloud season, STAC limits, and OSM completeness affect inputs.  
- **Models:** LULC and CNN performance vary by epoch and city; report `accuracy_report.csv` and `cnn_accuracy_report.csv` alongside claims.  
- **Scale:** Phase 6 drivers are not native-resolution surfaces; refine `coarsen_fact` only if compute allows.  
- **Statistics:** Imbalanced growth samples, residual spatial dependence, and single realizations of train/test splits for the CNN warrant sensitivity runs.  
- **Future:** Independent validation, causal designs (e.g. matching), uncertainty layers, and integration with socio-economic covariates where licensed.

<hr />

# Conclusions

This document bundles **what the pipeline produced** under explicit configuration, separates **deliverables** from **non-claims**, and lists **research gaps** so the work can be cited or extended honestly. **Primary archival format:** `outputs/urban_sprawl_report.pdf` (rendered by Phase 7 when LaTeX is available). For interactive mapping, open **`outputs/urban_sprawl_dashboard.html`**. For line-by-line technical detail, see **`README.md`** and **`docs/PROJECT_IMPLEMENTATION.txt`**.

<hr />

# References (illustrative)

1. Microsoft Planetary Computer STAC API and signed cloud-optimized GeoTIFF access.  
2. Hafner, S., et al. **landscapemetrics** — R package for landscape ecology metrics.  
3. McGarigal, K. et al. FRAGSTATS and landscape pattern metrics (conceptual background).  
4. Literature on Siamese / metric-learning architectures for remote-sensing change detection (survey papers, 2018–present).  
5. OSM contributors; WorldPop / DEM where used in optional Phase 6 layers.
